\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{array}
\usepackage{booktabs}
\usepackage[
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue
]{hyperref}

\title{Learning Signal Density: External-Sample and Internal-Compute Efficiency in Learning Loops}
\author{Francesco Giannicola \\ Limes Labs}
\date{June 2026}

\newcommand{\LSD}{\mathrm{LSD}}
\newcommand{\E}{\mathbb{E}}

\begin{document}
\maketitle

\begin{abstract}
This draft studies whether apparent sample efficiency should be measured at the
level of a whole learning loop rather than at the level of a model update
alone. We define learning signal density as heldout quality improvement per
external observation and charged internal compute, then evaluate raw examples,
question-answer expansion, oracle counterfactual expansion, and train-only
ranked induction policies in a controlled causal-text domain. The present
evidence is intentionally narrow: deterministic CPU tiny-MLP artifacts with
fresh-seed confirmation, explicit heldout isolation, and generated tables
pulled from committed JSON. The strongest current positive result is not a
frontier-model claim. It is a reusable measurement protocol plus a concrete
finding: representation width and model profile can improve density, but
scarce-sample generated-label failures remain visible at 24 and 32 external
materials; a hidden-rulebook audit shows that label correctness alone does not
explain those failures, and even a non-deployable post-hoc policy envelope
reaches the current target only at 48 materials. A deployable validation
coverage proxy turns part of that diagnostic into positive fresh-seed evidence
at 32 materials, but also exposes a 24-material failure, so the selector
problem remains open. A coverage-prior selector fixes that 24-material
over-selection and cuts coverage-selection cost, but still does not beat the
cheap train-size gate on density. A train-only tempered-budget ablation reduces
the sample-aware 24--32 material loss, but still does not beat raw-text
abstention at those scarce budgets. A compact train-size gate keeps that
abstention schedule, then improves the 64-material density point by dropping
redundant original QA examples only at the large-sample tier. A diversity
interaction probe improves the 64-material sample-aware row but does not beat
the compact density frontier, separating candidate-order coverage from
representation cost. A high-budget density cap then shows the frontier can
switch back to raw text once external evidence is abundant, trading absolute
gain for better learning-signal density. Support-ramped, fixed-window,
train-support-density, reuse-aware support-probe, validation-calibrated
support-precision, no-window gate, and transfer-stress follow-ups expose the
remaining selector problem: local transition windows can improve boundary rows,
and cheap validation calibration can improve one high-budget frontier, but late
false positives, seed-block transfer, transition-prior dependence, and
inspection cost still prevent a solved adaptive policy.
\end{abstract}

\paragraph{Keywords.}
sample efficiency; data transformation; synthetic data; counterfactual
generation; internal compute; learning-loop measurement; small neural audits.

\section{Introduction}

Token-count comparisons between biological learning and language-model
pretraining often treat data as if it arrived in a single undifferentiated
stream. That framing misses the mechanism this work wants to measure. A
learning system selects observations, rewrites them, asks questions, generates
counterfactuals, replays fragments, receives feedback, and spends internal
compute before any parameter update becomes useful.
Nearby mechanisms include token/data selection
\cite{lin2024rho1,li2024datacomp,xia2024less}, process feedback
\cite{lightman2023verify}, and replay \cite{schaul2015prioritized}. This draft
does not treat those literatures as interchangeable; it asks how to charge and
compare them inside one loop-level measurement protocol.

The research question is therefore not ``are language models or brains more
sample efficient?'' The sharper question is:
\emph{how much heldout improvement does a learning loop obtain per external
event once transformation, replay, feedback, selection, and internal compute
are charged?} This draft begins with a small but fully auditable causal domain
because the hidden rulebook, train/validation/heldout split, and generated-label
provenance can be controlled exactly.

\paragraph{Contributions in the current repository.}
The repository currently provides four pieces of evidence. First, it defines a
costed artifact schema for external events, internal examples, internal tokens,
charged transform cost, neural step count, estimated neural operations, and
claim scope. Second, it implements train-only and oracle transformation
policies under fixed heldout isolation. Third, it records fresh-seed tiny-MLP
sample-budget, feature-width, and epoch/width ablations. Fourth, the paper
tables below are generated from committed result JSON, so the manuscript cannot
silently drift away from the artifacts.

\paragraph{Non-claims.}
This is not a transformer-scale result, not a language-model leaderboard, and
not evidence that synthetic data is universally good. The current findings are
mechanism evidence for a measurement protocol. They justify more serious
replications; they do not replace them.

\section{Measurement Protocol}

Let a learning loop be
\[
S = (\theta, D_{\mathrm{ext}}, \pi_{\mathrm{select}},
\pi_{\mathrm{transform}}, \pi_{\mathrm{train}}, R, C).
\]
Here $\theta$ is the learner, $D_{\mathrm{ext}}$ is the externally observed
material, $\pi_{\mathrm{select}}$ chooses what to use, $\pi_{\mathrm{transform}}$
creates internal training examples, $\pi_{\mathrm{train}}$ updates the learner,
$R$ is the feedback or evaluation channel, and $C$ is charged internal compute.

For a heldout quality metric $Q$, external-event count $N_{\mathrm{ext}}$, and
charged internal compute $C_{\mathrm{int}}$, we track
\[
\mathrm{external\ sample\ efficiency} =
\frac{\Delta Q}{N_{\mathrm{ext}}},
\qquad
\mathrm{compute\ efficiency} =
\frac{\Delta Q}{C_{\mathrm{int}}},
\]
and
\[
\LSD =
\frac{\Delta Q}{N_{\mathrm{ext}} C_{\mathrm{int}}}.
\]
The reported signed version uses improvement over the majority baseline as
$\Delta Q$. Negative values remain negative instead of being clipped away.

\section{Anti-Cheat Rules}

All transformation and selection policies must be chosen without heldout
examples. Validation labels may be used only by conditions that declare and
charge validation use. Train-only ranked induction may use confidence, support,
and salience signals computed from training material, but it may not inspect
heldout accuracy. Oracle counterfactual expansion is allowed only as an upper
bound inside the synthetic world and must be labeled as oracle-generated.

Every committed result artifact carries a claim scope. The current neural
artifacts explicitly record that they are fresh-seed confirmations, use a
synthetic domain, do not use heldout data for selection, and are not yet
paper-ready claims. The generated paper-table script refuses artifacts that
violate those scope flags.

\section{Experimental Design}

The domain is a synthetic causal-text world with hidden rules and fixed splits
by material/stimulus pair. The learner observes textual or question-answer
views of a material, predicts heldout labels, and reports improvement over a
majority baseline. The current neural backend is a deterministic CPU MLP with
hashed features. It is intentionally dependency-light so that the artifact can
run in ordinary CI, but it is still a neural replication rather than the first
linear pilot.

The latest ablation path asks two practical efficiency questions. First, once
1024 hashed features improve the signal frontier, how much epoch/width budget
is actually needed? The tested profiles compare 32x8, 16x8, and 8x8 MLPs at
1024 features over 16, 24, 32, 48, and 64 external-material budgets. Second,
does explicit reliability selection repair the scarce-sample generated-label
failure? To test that, the 16x8 profile is rerun with a train-only agreement
gate, charged validation-ranked induction, and MDL rule expansion beside the
train-only ranked policies and oracle counterfactual headroom.
The latest selector extensions replace the expensive full validation portfolio
with a two-epoch linear proxy that scores the same candidate policies before
training only one final MLP, then test whether a raw-text abstention margin
reduces scarce-budget downside. A final train-size gated baseline removes
validation selection entirely: it uses raw text below 144 train events and
sample-aware self-ranked induction only after the train split is large enough.
The latest diagnostics then audit generated synthetic labels against the hidden
rulebook and heldout motif distribution after the selector-transfer sweep has
run. These audits are non-deployable by construction, but they separate label
correctness and coverage from downstream neural gain.
The latest deployable follow-up replaces heldout coverage with validation motif
coverage: it scores candidate generated-label policies by motif-distribution
fit to validation, does not use validation labels for the selector score, and
charges candidate construction plus validation-motif scanning before training
one final tiny MLP.
The latest coverage-prior control then adds a 96-train-event raw-text floor,
prunes the coverage scan to a lean candidate set, and adds a small charged
compute penalty to the validation-motif coverage score.
The latest train-only follow-up then lowers the sample-aware mid-budget
synthetic ratio from 0.75 to 0.50 to test whether fewer generated labels repair
the same failure without validation selection.
The latest efficiency follow-up keeps the train-size abstention schedule but
changes the large-sample representation: above 224 train events it retains raw
original observations and ranked generated examples, while dropping original
QA duplicates.
A diversity interaction follow-up then keeps the same train-only budget and
split discipline while changing only ranked-candidate ordering. It compares
plain diversity, sample-aware diversity, and compact diversity to separate
coverage balancing from representation-cost savings.
A high-budget extension then tests a density-capped schedule through 128
materials: it keeps compact induction through 96 materials and returns to raw
text once the train split reaches the abundant-data tier.
A second high-budget extension tests the middle of that gain-density tradeoff:
it keeps compact induction but raises the induced-label minimum support from 3
to 4 after the abundant-data tier.
A final high-budget control then asks whether tightening the induced-label
confidence floor after 432 train events is enough to repair the remaining
tradeoff.
A fixed-window high-budget control then asks whether simple train-event
breakpoints are enough: compact below 320 train events, raw from 320 to 400,
support-ramped compact from 400 to 432, and raw again after 432.

\section{Results}

Table~\ref{tab:f1024-profile-frontier},
Table~\ref{tab:validation-selected-reliability-probe},
Table~\ref{tab:post-hoc-policy-envelope},
Table~\ref{tab:validation-portfolio-selector},
Table~\ref{tab:selector-transfer-stress},
Table~\ref{tab:generated-label-audit},
Table~\ref{tab:generated-coverage-audit},
Table~\ref{tab:validation-coverage-proxy},
Table~\ref{tab:validation-coverage-prior},
Table~\ref{tab:tempered-sample-aware},
Table~\ref{tab:train-size-gated-baseline},
Table~\ref{tab:compact-train-size-gated},
Table~\ref{tab:diversity-interaction},
Table~\ref{tab:density-capped-compact},
Table~\ref{tab:support-ramped-compact},
Table~\ref{tab:late-confidence-ramped-compact},
Table~\ref{tab:density-window-compact},
Table~\ref{tab:train-support-density-selector},
Table~\ref{tab:support-probe-window-selector},
Table~\ref{tab:validation-support-precision-selector},
Table~\ref{tab:validation-support-precision-gate},
Table~\ref{tab:support-selector-transfer}, and
Table~\ref{tab:low-budget-failure-probe} are generated from the checked-in JSON
artifacts by the paper-table generator at
\texttt{scripts/build\_\allowbreak paper\_\allowbreak tables.py}.

\input{generated/results_tables.tex}

The current frontier has two useful signals and one important failure. First,
the 16x8 profile halves neural training operations relative to 32x8 for the
same feature width and improves the best self-ranked signed gain, so larger
training budgets are not automatically better. Second, reducing the profile to
8x8 halves operations again, but loses the self-ranked target and weakens the
main non-oracle result. Third, scarce-sample generated-label noise is not fixed
by shrinking the optimizer budget: all ranked 16x8 and 8x8 rows still fail at
32 materials, even when sample-aware ranking is positive at 16 and 24 for the
8x8 profile.

The reliability-policy rerun sharpens that failure mode. A train-only agreement
gate, which keeps only induced labels supported by agreeing rule projections,
does not solve the problem: it is worse than sample-aware ranking at 16
materials, tied with it at 32, and has much lower peak signed gain. Charged
validation-ranked induction slightly improves the 48-material point relative to
self-ranked induction, but it does not move the first target below 48 and it
does not repair the 32-material failure. MDL rule expansion is more conservative
at scarce budgets: it is positive at 16 materials and less negative than
self-ranked induction at 32 materials. However, it still remains below target at
32, reaches target only at 48, and has a lower peak signed gain than the
train-only ranked policies. This points to a more specific next step: a useful
generated-label policy must improve reliability at 24--32 materials without
paying so much selection cost that density collapses.

The post-hoc envelope in Table~\ref{tab:post-hoc-policy-envelope} is
deliberately not a deployable method: it chooses the best non-oracle condition
at each budget after seeing the completed heldout results. That diagnostic
still does not reach the signed-gain target below 48 materials. At 24 materials,
raw text is the best non-oracle row because every generated-label policy hurts
heldout accuracy. At 32 materials, MDL rule expansion is only the least bad
choice, not a positive result. The envelope switches again to validation-ranked
induction at 48 and self-ranked induction at 64. This makes the next research
problem sharper: a useful policy must learn when to abstain, when to compress
rules, and when to rank generated labels using only train signals or explicitly
charged validation signals.

Table~\ref{tab:validation-portfolio-selector} turns that diagnostic into three
deployable selector tests. The full portfolio trains six non-oracle candidate
MLPs, evaluates them on validation examples, chooses by validation improvement
over majority, charges the whole search, and then evaluates heldout once. The
linear proxy selector scores the same candidate policies with a two-epoch
linear classifier, charges those proxy fits, trains only the selected tiny MLP,
and then opens heldout once. The abstaining proxy adds a raw-text fallback: a
non-raw candidate must beat raw text by three validation examples before the
selector leaves raw text. The plain proxy improves the deployable selector
frontier, but it is still not a solved policy. At 48 materials it reaches 0.103
signed gain versus 0.076 for the full portfolio; at 64 materials it reaches
0.153 signed gain versus 0.109, while improving charged signed LSD from
0.000627 to 0.002091. The abstaining proxy improves downside control at scarce
budgets: it moves 16 materials from -0.105 to 0.000 signed gain and 24
materials from -0.041 to 0.006897. However, it still remains negative at 32
materials, lowers the 48-material gain to 0.082759, first reaches the target
only at 48, and still has lower 64-material density than fixed self-ranked and
sample-aware self-ranked induction. The result is therefore a real
selector-risk improvement, not a breakthrough on scarce-sample reliability.

Table~\ref{tab:selector-transfer-stress} then asks whether that selector story
survives a fresh-seed stress test on seeds 37, 41, 43, 47, and 53. It mostly
does not. The abstaining and linear proxy selectors still first reach target at
48 materials, but their 64-material signed gain falls from 0.153247 on the
development selector artifact to 0.114286 on fresh seeds. The full portfolio
selector reaches 0.127273 at 64 materials, but its charged density remains far
below fixed sample-aware self-ranked induction. At 32 materials, all deployable
selector variants remain negative, and raw text is less negative than the
selector family. This is useful negative evidence: the current validation
signals are not stable enough to drive a deployable adaptive policy.

Table~\ref{tab:generated-label-audit} narrows that reliability problem. At 32
materials, agreement-gated generated labels are 0.917241 correct against the
hidden rulebook, but their linked neural gain is still -0.131579; sample-aware
labels are less precise at 0.675862 and also lose more at -0.173684. Oracle
counterfactual labels are perfectly correct and yet reach only 0.015790 signed
gain at the same budget. At 64 materials the contrast is sharper:
agreement-gated labels are more precise than sample-aware labels, 0.954783
versus 0.829565, but produce lower gain, 0.080519 versus 0.142857. This means
the scarce-budget failure is not reducible to generated-label correctness.
Coverage, distribution shift, interaction with the tiny learner, and charged
selection cost remain live mechanisms.

Table~\ref{tab:generated-coverage-audit} makes the distribution part concrete.
The table compares generated synthetic-label motif frequencies with heldout
family/stimulus/modifier frequencies after the sweep has run. Lower triple L1
distance means the generated labels land in a motif distribution closer to
heldout. At 64 materials, sample-aware induction has lower triple L1 distance
than agreement-gated induction, 0.477843 versus 0.576443, and higher gain,
0.142857 versus 0.080519, despite lower label precision. At 32 materials,
validation-ranked induction is much less negative than agreement-gated
induction and also has lower triple L1 distance, 0.683666 versus 0.764428.
This does not make heldout distribution a usable selector signal; it identifies
what a future train/validation-side proxy must approximate.

Table~\ref{tab:validation-coverage-proxy} is the first deployable version of
that approximation. It uses validation motif distribution, not heldout
distribution or validation labels, to score the same candidate family. On fresh
seeds 103, 107, 109, 113, and 127, the proxy moves the 32-material row from
negative fixed sample-aware induction to positive signed gain, 0.010526, by
abstaining to raw text on four seeds and choosing MDL once. It also reaches
0.171428 signed gain at 64 materials, above the abstaining proxy but below
fixed sample-aware induction and the validation portfolio in this confirmation
artifact. The failure is equally important: at 24 materials it
selects MDL rule expansion on all five seeds and falls to -0.082759, worse than
raw text and the train-size gate. The mechanism signal is therefore real enough
to keep, but the current rule needs a low-budget abstention floor or train-size
prior before it can be promoted.

Table~\ref{tab:validation-coverage-prior} tests that floor directly on fresh
seeds 601, 607, 613, 617, and 619. The selector uses raw text below 96 train
events; above that floor it scans only raw text, sample-aware self-ranked
induction, and validation-ranked induction, scoring generated-label candidates
by validation-motif coverage plus a small charged-compute penalty. This removes
the 24-material coverage-proxy failure, moving the row from -0.062069 to
0.000000, and improves the coverage-proxy signed LSD at 48 and 64 materials
from 0.003147 and 0.003171 to 0.004756 and 0.005001. The negative result is
still decisive: the train-size gate reaches the same 48/64 gains with higher
signed LSD, 0.005552 and 0.006089, because it avoids selector scanning
altogether. Coverage-prior selection is therefore cost control, not yet a
deployable selector frontier.

Table~\ref{tab:tempered-sample-aware} tests whether that floor can be replaced
by a milder train-only budget change. The tempered policy keeps the
sample-aware support and confidence rule, but lowers the synthetic budget ratio
from 0.75 to 0.50 for train splits below 144 events. On fresh seeds 157, 163,
167, 173, and 179, this reduces sample-aware damage at 24 materials,
-0.096552 versus -0.137931, and at 32 materials, -0.078947 versus -0.152632,
while preserving the 48- and 64-material gains. The negative part is again the
important part: raw text and the train-size gate are still better at 24 and 32
materials. Smaller generated-label budgets help, but scarce-budget abstention
is still a stronger baseline than tempered synthesis.

Table~\ref{tab:train-size-gated-baseline} checks a simpler alternative on a
second unseen seed set: use raw text below 144 train events, then switch to
sample-aware self-ranked induction. This train-only schedule reaches the target
at 48 materials and matches fixed sample-aware induction at 64 materials
(0.145454 signed gain and 0.005090 signed LSD), without validation selector
search. It is not a scarce-budget solution: it is still negative at 16 and 32
materials, and the validation portfolio is less negative at 32. Its value is
instead as a cheap promotion gate. Any future adaptive selector should justify
its extra selection cost by beating this schedule on fresh seeds.

Table~\ref{tab:compact-train-size-gated} asks whether the high-budget part of
that schedule can be made denser without changing the generated labels. The
compact condition is identical to the train-size gate through 48 materials. At
64 materials, on fresh seeds 181, 191, 193, 197, and 199, it drops original QA
duplicates while keeping raw originals and ranked generated examples. Charged
compute falls from 124274.8 to 77354.8 units, signed gain rises from 0.132467
to 0.140260, and signed LSD rises from 0.004634 to 0.007883. This is a useful
efficiency result, not a scarce-budget solution: it does not change the 16--32
material rows, and it relies on the same train-only abstention schedule.

Table~\ref{tab:diversity-interaction} asks whether balancing generated-label
coverage should be treated as a candidate-ordering improvement or as a
replacement for compact representation savings. On fresh seeds 701, 709, 719,
727, and 733, plain diversity improves self-ranked gain at 24, 32, and 64
materials. The sample-aware diversity variant improves the 64-material
sample-aware row from 0.158441 gain and 0.005544 signed LSD to 0.168831 and
0.005908. The compact-diversity variant is the important negative control: it
keeps the low-budget compact behavior but lowers the 64-material compact row
from 0.135065 gain and 0.007594 signed LSD to 0.116883 and 0.006573.
Diversity is therefore a useful train-only ordering signal for the fuller
sample-aware representation, but it does not replace the density benefit from
dropping redundant original QA views.

Table~\ref{tab:density-capped-compact} extends the question beyond the original
64-material frontier. The density-capped policy keeps compact induction through
96 materials, then returns to raw text from 104 materials onward. On fresh
seeds 293, 307, 311, 313, and 317, this keeps the compact 64- and 80-material
density gains and exposes a high-budget tradeoff. At 128 materials, compact
induction has higher signed gain than the density cap, 0.132468 versus
0.081818, but the density cap has higher signed LSD, 0.003452 versus 0.001860,
because it avoids the generated-label transform cost. The conclusion is not
that raw text dominates. It is that a useful policy should choose between raw,
compact, and fuller generated-label views according to the gain-density
frontier, not only according to heldout accuracy.

Table~\ref{tab:support-ramped-compact} tests a middle policy on fresh seeds
401, 409, 419, 421, and 431. The support-ramped condition matches compact
induction through 96 materials, then requires support 4 instead of support 3
for induced labels. The result is a Pareto tradeoff, not a universal
improvement. At 104 materials it improves signed LSD over both plain compact
and raw fallback, 0.003735 versus 0.002857 and 0.002145, while keeping higher
gain than raw. At 128 materials it recovers more signed gain than both raw
fallback and plain compact, 0.184416 versus 0.154545 and 0.119481, but raw
fallback remains denser, 0.006521 versus 0.005153. This makes the high-budget
selector target more precise: the choice is not just synthesize or abstain, but
which support floor buys the desired point on the gain-density frontier.

Table~\ref{tab:late-confidence-ramped-compact} is a negative control for the
same idea. It keeps the support-ramped policy, but after 432 train events it
raises the induced-label confidence floor from 0.55 to 0.60. On fresh seeds
499, 503, 509, 521, and 523, this improves the 120-material support-ramped row
from 0.134722 gain and 0.004119 signed LSD to 0.138889 and 0.004322. The
improvement is real but not enough: raw/density-capped fallback is still better
at the same row, 0.145833 gain and 0.006977 signed LSD, and the
late-confidence policy's best signed gain, 0.171098, remains below plain
compact induction's 0.196875. A confidence floor alone is therefore not the
missing selector.

Table~\ref{tab:density-window-compact} tests whether a fixed transition window
can encode the high-budget selector. On fresh seeds 929, 937, 941, 947, and
953, the density-window policy matches compact at 64 and 80 materials, raw at
96, 104, and 120, support-ramped compact at 112, and raw again at 128. This is
locally useful: at 112 materials it improves signed LSD over density-capped
raw fallback, 0.004269 versus 0.004001, and at 120 materials it keeps raw's
0.005648 signed LSD instead of the support-ramped 0.004899. It also fails in a
way that matters: at 128 materials it falls back to raw at 0.003726 signed
LSD, while support-ramped compact reaches 0.004290. Fixed train-event windows
are therefore a good control, but not a deployable selector.

Table~\ref{tab:train-support-density-selector} then tests a slightly more
adaptive train-only selector. It chooses among raw text, compact train-size
gated induction, and support-ramped compact induction using support kept per
charged compute, and it charges candidate inspection before the final fit. On
fresh seeds 1033, 1039, 1049, 1051, and 1061, the selector chooses
support-ramped compact at 104 and 112 materials, raw text for every
128-material run, and a raw/support mix at 120. Those choices are sensible, but
the measurement is not free: at 104 materials the selector matches
support-ramped gain while lowering signed LSD from 0.003994 to 0.003262; at 120
materials it reaches 0.003844 signed LSD while raw and density-window reach
0.005515; at 128 materials it reaches 0.004142 while raw reaches 0.005809. The
result is a selector-cost failure that points to cheaper proxies or
value-of-information gates, not a new frontier policy.

Table~\ref{tab:support-probe-window-selector} tests that value-of-information
idea directly. The policy uses compact induction below 320 train events, raw
text outside the 360--432 support-probe window, and inside that window inspects
only support-ramped compact induction; if support is selected, the inspected
candidate construction is reused. On fresh seeds 1063, 1069, 1087, 1091, and
1093, this removes the 104-material overhead: the selector reaches the
support-ramped row at 0.004655 signed LSD instead of the full
train-support-density selector's 0.003805. The result is still not solved. At
112 materials, it chooses support while raw/density-capped reaches 0.007103
signed LSD versus 0.005161; at 120 materials, the no-probe raw fallback reaches
0.003588 while support-ramped compact reaches 0.004240. Accounting reuse helps,
but the train-only policy signal is still too coarse.

Table~\ref{tab:validation-support-precision-selector} adds a cheap validation
calibration to the same raw/support boundary. The selector keeps compact below
320 train events, keeps the fixed 400--432 support transition, and otherwise
uses validation labels only to estimate eligible induced-prediction precision
before choosing support-ramped compact or raw text. On fresh seeds 1259, 1277,
1279, 1283, and 1289, it improves the 96-material signed LSD from the
support-probe/raw row's 0.002075 to 0.004198 and the 104-material row from
0.004331 to 0.004455. Averaged across 64--128 materials, it reaches 0.006138
signed LSD versus 0.005941 for the support-probe window. The result remains
mixed rather than final: at 120 materials raw/support-probe reaches 0.006246
while the validation selector reaches 0.005723, and at 128 raw reaches 0.005206
while the selector reaches 0.004866.

Table~\ref{tab:validation-support-precision-gate} asks whether the fixed
400--432 support transition is still needed once validation precision is
available. The no-window gate applies the same precision threshold everywhere
above the compact floor. On fresh seeds 1381, 1399, 1409, 1423, and 1427, this
slightly beats the support-probe average, 0.006104 versus 0.006074 signed LSD,
but it falls below the fixed-transition validation selector at 0.006223. The
clearest miss is 112 materials: the no-window gate reaches 0.005031 signed LSD,
while the fixed-transition selector and support-ramped compact reach 0.005864.
The fixed transition prior is therefore carrying real information in this
frontier, even though it is not a final adaptive solution.

Table~\ref{tab:support-selector-transfer} reruns the support-selector family on
fresh seeds 1459, 1471, 1481, 1483, and 1487 after the fixed-transition selector
and no-window gate have both been developed. The no-window gate transfers better
than the fixed-transition validation selector on average, 0.005936 versus
0.005601 signed LSD, and it partially repairs the 112-material fixed-prior miss
by reaching 0.006517 instead of 0.004175. The result is still a negative transfer
stress for selector promotion: the simple train-only density-capped raw fallback
has the strongest all-budget average at 0.006115, with especially visible
advantages at 112, 120, and 128 materials. Thus, neither fixed transition priors
nor validation precision alone are stable enough to replace an explicit
expected-utility model.

\section{Discussion}

The mathematical shape suggested by the current evidence is a four-way
frontier. Feature width controls collision and representation capacity. Epoch
and hidden-unit budget control how much of that representation the learner can
use. Transformation policy controls whether generated labels add reliable
signal or amplify noise when external material is scarce. Selection policy
controls how much reliability is bought per extra unit of charged compute. A
useful next breakthrough is therefore unlikely to come from only increasing or
shrinking the MLP budget. The low-budget failure points instead toward
reliability-aware generated-label policies with sharper uncertainty estimates,
or a small transformer backend whose sequence structure can be charged honestly.
The post-hoc envelope also turns the problem into a selector-learning question:
future experiments should preregister train/validation features that predict
whether raw text, compact rules, validation-ranked candidates, or train-only
ranked candidates are appropriate before heldout evaluation is opened.
The selector probes are useful controls for that plan: the full portfolio shows
that trying every candidate on validation is too expensive, the linear proxy
shows that cheaper model hierarchies help, and the abstaining proxy shows that
simple validation-count caution can reduce downside at 16--24 materials without
repairing the 32-material failure. The fresh-seed transfer test adds a stronger
constraint: future selectors must beat simple fixed policies on new seeds, not
only on the artifacts that inspired them. The train-size gated baseline makes
that constraint sharper: selector search has to buy more than a cheap
train-size schedule that abstains from generated labels at scarce budgets.
The generated-label audit adds one more constraint: a policy that optimizes
only synthetic-label precision can still lose if the labels cover the wrong
regions or interact poorly with the learner.
The generated-coverage audit suggests one measurable version of that warning:
heldout-like motif coverage appears closer to the gain ordering than raw label
precision in the selector-transfer artifact, but it must be approximated
without heldout access before it can become a policy.
The validation-coverage proxy is a first approximation, and its mixed result
sharpens the next design constraint: coverage scoring needs to be joined with
scarce-budget abstention, because distributional fit can still choose a
generated-label policy when raw text is the better action.
The coverage-prior selector confirms that this repair is possible but
incomplete: a train-size floor and lean candidate scan remove one over-selection
failure and reduce charged cost, yet the selector still has to justify its
validation-motif scan against a no-scan train-size schedule.
The tempered sample-aware ablation sharpens the same point from the other
direction: reducing generated-label count lowers damage, but the better
decision at 24--32 materials is often not fewer synthetic examples; it is no
synthetic examples yet.
The compact train-size gate adds the large-sample complement: once generated
labels are useful, retaining every original representation is also a policy
choice, and dropping redundant original QA views can improve density without
changing the train-only generated labels.
The diversity interaction probe separates representation cost from candidate
ordering: diversity can improve sample-aware accuracy, but it is not enough to
beat compact density after redundant original views are removed.
The density-capped compact probe adds the abundant-data complement: even
generated labels that improve accuracy can become density-inefficient once raw
external evidence is plentiful enough. The next selector therefore needs a
utility model over both gain and charged compute, not just a classifier for
``should synthesize''.
The support-ramped compact probe adds another constraint to that utility model:
increasing induced-label support can recover high-budget gain-density tradeoffs
that raw fallback discards, but it still does not dominate raw on density
alone.
The late-confidence control further narrows the design: stricter generated
labels can improve one support-ramped point, but confidence tightening alone
does not produce a deployable frontier policy.
The density-window compact control narrows it again: simple train-event
breakpoints can recover the 112-material transition and avoid unnecessary
support-ramped cost at 120, but they miss the 128-material support-ramped
opportunity. A useful selector likely needs a sample-aware utility signal over
expected gain, representation cost, and generated-label reliability.
The train-support-density selector makes that requirement sharper: even a
train-only signal that chooses plausible high-budget representations can lose
once candidate inspection is charged. The next useful policy should decide not
only which representation is best, but whether the expected value of inspecting
that representation exceeds the cost of the inspection itself.
The support-probe window control answers part of that question by reusing
selected candidate construction, but it leaves a policy-quality failure: fixed
train-event windows still choose the wrong side of the raw/support tradeoff on
some fresh seed rows. A stronger selector needs train-only evidence about
expected heldout gain, not only support density and event-count windows.
The validation support-precision selector shows that a cheap validation signal
can improve the boundary rows and the average high-budget density frontier, but
it also creates late false positives once support-ramped compact is no longer
consistently denser than raw text. The next selector should combine reliability
calibration with an explicit expected-cost term rather than treating high
precision as sufficient evidence to synthesize.
The no-window gate narrows that lesson: validation precision alone does not
replace the transition prior. It can avoid some late support choices, but it also
gives up useful support at the 112-material transition. A deployable selector
needs a utility model that combines validation reliability, train-event regime,
and representation cost instead of depending on any one of those signals alone.
The support-selector transfer stress then turns that lesson into a promotion
gate. A fixed transition can help one seed block and harm the next; a validation
gate can improve over that fixed prior and still lose to a no-scan density cap.
The post-hoc error audit in Table~\ref{tab:support-selector-error-audit}
quantifies that failure mode across the support-selector chain: the
least-regret transfer selector is still positive-regret against the best simple
comparator (\(0.000496\) signed LSD), and it beats that comparator on only
\(1/7\) budgets. The one local negative-regret result appears on the development
gate artifact, not on the transfer block, so it should be treated as a design
clue rather than a promotion signal.
The next selector should be evaluated as an expected-value-of-information
problem: inspect or validate only when the anticipated gain over the raw or
density-capped fallback exceeds the charged cost of obtaining that evidence.

The oracle counterfactual condition remains valuable because it estimates the
headroom available if generated examples were reliable. It should not be
confused with a deployable method. The non-oracle target is to close part of
the gap with train-only or explicitly charged validation mechanisms.

\section{Release Readiness}

This draft follows the release discipline used in Limes Labs public research
repos, but it is not ready for a versioned paper release. A release-quality
package should include a rendered PDF, source archive, result manifest,
checksums, artifact README, submission metadata, and the exact commands used to
rebuild the paper and reproduce the canonical artifacts. The current repo has
the result artifacts and generated table path; it still needs stronger model
replication and a frozen release bundle before public paper promotion.

\section{Limitations}

The current neural backend is a tiny MLP, not a transformer or language model.
The domain is synthetic and deliberately narrow. Oracle counterfactual labels
use the hidden world and exist only as an upper bound. The sample-aware and
self-ranked policies were developed through earlier discovery artifacts, so
fresh-seed confirmation helps but does not eliminate the need for independent
replication. Wall-clock time and hardware metadata are not yet sufficient for
a full compute-efficiency claim.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
